Automatically proving equivalence by type-safe reflection

### 2016-01-18 paper-JFP-DTP/content/1-introduction.tex
Added bibtex

Proving that one term is equal to another one within some theories is something quite common in formal certification. The two terms are usually not syntactically equal --this case would be trivial--, but they are equal according to some properties. These properties are either obtained by implementation (the implementation of $+$ for natural numbers satisfies the property of associativity for example), or by axioms if we are working within an axiomatised structure.

For example, when working with natural numbers, one might need to prove the lemma $\forall\ xy:Nat,\ x*x\ +\ 3*x\ +\ x*y\ =\ x*(3+y+x)$, and when working with lists and concatenation of lists, one might need to prove something like $\forall\ l1\ l2\ l3:List A,\ ((l1\ ++\ (l2 ++ l3))\ ++\ l4\ =\ (l1++l2)\ ++\ (l3++l4)$. Proving this kind of auxiliary lemmas is time consuming, is much too specific for being reusable, and isn't the interesting part of formal certification. Therefore, we would like to be able to automatize a far as possible the constructions of this kind of proofs for the Idris language, a dependently typed language which doesn't contain many proof automations at the moment. \\
\\
When doing formal certification, many functions have a lemma of correctness that talks about equality of terms. More precisely, the following situation is very familiar. The function that we wants to certify is $f\ :\ A \rightarrow B$. We have two functions $sem_A\ :\ A \rightarrow SEM$ and $sem_B\ : \ B \rightarrow SEM$ of interpretation which associate a semantic to each type A and B. If this semantic says everything that matters for the behaviour of $f$, then the lemma of correctness of f can be expressed as $forall\ a:A,\ semA\ a = semB (f\ a)$. The interesting thing is that, very often, the type $SEM$ has the structure of some nice algebraic structure, and it becomes thus possible to decide equalities in it automatically.
All the equality symbols we have encountered so far refer to the propositional equality. We remind the readers in the next subsection how this equality is defined in Idris, and more generally in Intentional Type Theories, and how it differs from the --more primitive-- judgemental equality.

>>>>

Proving that one term is equal to another within some theories is quite common
in formal certification. The two terms are usually not syntactically 
equal---this case would be trivial---but they are equal relative to some
equivalence properties. These properties are either obtained by lemmas
(the implementation of $+$ for natural numbers satisfies the property of
associativity for example), or by axioms.

For example, when working with natural numbers, one might need to prove the
lemma $\forall\ x, y:Nat,\ x*x\ +\ 3*x\ +\ x*y\ =\ x*(3+y+x)$. Or,
when working
with lists and concatenation of lists, one might need to prove the lemma
$\forall\ l1,\ l2,\ l3:List A,\ ((l1\ ++\ (l2 ++ l3))\ ++\ l4\ =\ (l1++l2)\ ++\
(l3++l4)$. Proving this kind of auxiliary lemmas is time consuming,
typically not reusable due to being very specific, and is rarely
the interesting part of formal
certification. Therefore, we would like to be able to automatize as far as
possible the constructions of this kind of proofs. We have implemented
the machinery to do so in Idris, a
dependently typed language~\cite{brady2013idris}.

When doing formal certification, many functions have a lemma of correctness that talks about equality of terms. More precisely, the following situation is very familiar. The function that we wants to certify is $f\ :\ A \rightarrow B$. We have two functions $sem_A\ :\ A \rightarrow SEM$ and $sem_B\ : \ B \rightarrow SEM$ of interpretation which associate a semantic to each type A and B. If this semantic says everything that matters for the behaviour of $f$, then the lemma of correctness of f can be expressed as $forall\ a:A,\ semA\ a = semB (f\ a)$. The interesting thing is that, very often, the type $SEM$ has the structure of some nice algebraic structure, and it becomes thus possible to decide equalities in it automatically.
All the equality symbols we have encountered so far refer to the propositional equality. We remind the readers in the next subsection how this equality is defined in Idris, and more generally in Intentional Type Theories, and how it differs from the --more primitive-- judgemental equality.
